---
title: "02 Build the 10-min dataset and data availability (2024)"
author: "Francisca Muñoz Narbona"
format:
  html:
    toc: true
    self-contained: true
editor: visual
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
editor_options:
  chunk_output_type: console
---

Step 2 of the pipeline. Joins the processed station files of the Cerro Oyarbide transect (`data/stations/`, output of `01_merge_station_files.R`), the coastal airport reference station (`data/airport/`) and the GOES-16 fog and low cloud (FLC) detection (`data/goes/`) into a single 10-min dataset for 2024, applies the basic cleaning rules, computes the data availability table (Table 1 of the manuscript) and saves `data/derived/stations_10min_2024.csv`.

Variable and station names in the output keep the original (Spanish) column names used throughout the pipeline; see the data dictionary in the README.

```{r}
#| label: read-data

library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)

dir_data <- "../data"

oya518  <- read_csv(file.path(dir_data, "stations/oya_518_2024-2025.csv"))
oya780  <- read_csv(file.path(dir_data, "stations/oya_780_2024-2025.csv"))
oya862  <- read_csv(file.path(dir_data, "stations/oya_862_2024-2025.csv"))
oya1069 <- read_csv(file.path(dir_data, "stations/oya_1069_2024-2025.csv"))
oya1128 <- read_csv(file.path(dir_data, "stations/oya_1128_2024-2025.csv"))
oya1193 <- read_csv(file.path(dir_data, "stations/oya_1193_2024-2025.csv"))
oya1211 <- read_csv(file.path(dir_data, "stations/oya_1211_2024-2025.csv"))
oya1354 <- read_csv(file.path(dir_data, "stations/oya_1354_2024-2025.csv"))

# Coastal reference station (Diego Aracena airport, 48 m a.s.l.)
airport <- read_csv(file.path(dir_data, "airport/datos_2024_10min_aeropuerto.csv"))

# Airport wind speed (knots -> m/s); this record is hourly
airport_wind <- read_delim(file.path(dir_data, "airport/200006_2024_Viento_.csv"),
                           delim = ";") %>%
  select(Instante, ff) %>%
  rename(FechaHora_10min = Instante) %>%
  mutate(viento_vel_m_s = ff * 0.51444) %>%
  select(-ff)

airport <- airport %>%
  left_join(airport_wind, by = "FechaHora_10min")

# GOES-16 fog and low cloud (FLC) detection over station OYA1211
goes <- read_csv(file.path(dir_data, "goes/FLC_OYA1_2024.csv")) %>%
  rename(datetime = `...1`) %>%
  transmute(datetime = as.POSIXct(datetime, tz = "UTC"),
            FLC = FLC,
            estacion = "GOES") %>%
  pivot_longer(cols = FLC, names_to = "variable", values_to = "valor") %>%
  filter(year(datetime) == 2024)
```

```{r}
#| label: select-variables

# Manual selection and renaming of the variables used from each station

oya518_sel <- oya518 %>%
  transmute(
    datetime = datetime,
    temperatura = luft_temperatur_c,
    presion = luft_druck_h_pa,
    niebla_ml = niederschlag_nebelwasser_ml,
    pp_mm = niederschlag_menge_mm,
    estacion = "OYA_518",
    radiacion_w_m2 = strahlung_global_w_m2,
    humedad = luft_feuchte_percent,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    temp_sup_c = oberflaeche_temperatur_1_c
  )

oya780_sel <- oya780 %>%
  transmute(
    datetime = datetime,
    humedad = luft_feuchte_percent,
    temperatura = luft_temperatur_c,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    presion = luft_druck_h_pa,
    niebla_mm = niederschlag_nebelwasser_mm,
    estacion = "OYA_780"
  )

# fog water only in mL at this station (the mm record is dropped)
oya862_sel <- oya862 %>%
  transmute(
    datetime = datetime,
    temperatura = luft_temperatur_c,
    humedad = luft_feuchte_percent,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    niebla_ml = niederschlag_nebelwasser_ml,
    estacion = "OYA_862"
  )

# fog water only in mL at this station (the mm record is dropped)
oya1069_sel <- oya1069 %>%
  transmute(
    datetime = datetime,
    temperatura = luft_temperatur_c,
    humedad = luft_feuchte_percent,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    niebla_ml = niederschlag_nebelwasser_ml,
    estacion = "OYA_1069"
  )

# both fog water records (mm and mL) are kept
oya1128_sel <- oya1128 %>%
  transmute(
    datetime = datetime,
    temperatura = luft_temperatur_c,
    humedad = luft_feuchte_percent,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    niebla_mm = niederschlag_nebelwasser_mm,
    niebla_ml = niederschlag_nebelwasser_ml,
    estacion = "OYA_1128"
  )

# both fog water records (mm and mL) are kept
oya1193_sel <- oya1193 %>%
  transmute(
    datetime = datetime,
    temperatura = luft_temperatur_c,
    humedad = luft_feuchte_percent,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    niebla_mm = niederschlag_nebelwasser_mm,
    niebla_ml = niederschlag_nebelwasser_ml,
    estacion = "OYA_1193"
  )

# main station: fog and dew water (mm and mL), visibility, soil and radiation
oya1211_sel <- oya1211 %>%
  transmute(
    datetime = datetime,
    temperatura = luft_temperatur_c,
    humedad = luft_feuchte_percent,
    presion = luft_druck_h_pa,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    radiacion_w_m2 = strahlung_global_w_m2,
    temp_suelo_50cm_c = boden_temperatur_50cm_c,
    visibilidad = strahlung_sichtweite_m,
    temp_hum_10cm = boden_feuchte_10cm_percent,
    temp_suelo_10cm_c = boden_temperatur_10cm_c,
    temp_hum_20cm = boden_feuchte_20cm_percent,
    temp_suelo_20c_c = boden_temperatur_20cm_c,
    temp_sup_c = oberflaeche_temperatur_c,
    niebla_mm = niederschlag_nebelwasser_mm,
    niebla_ml = niederschlag_nebelwasser_ml,
    rocio_mm = niederschlag_tauwasser_mm,
    rocio_ml = niederschlag_tauwasser_ml,
    estacion = "OYA_1211"
  )

# both fog water records (mm and mL) are kept
oya1354_sel <- oya1354 %>%
  transmute(
    datetime = datetime,
    temperatura = luft_temperatur_c,
    humedad = luft_feuchte_percent,
    viento_vel_m_s = wind_geschwindigkeit_m_s,
    viento_dir = wind_richtung,
    niebla_mm = niederschlag_nebelwasser_mm,
    niebla_ml = niederschlag_nebelwasser_ml,
    estacion = "OYA_1354"
  )

airport_sel <- airport %>%
  transmute(
    datetime = FechaHora_10min,
    temperatura = ts,
    presion = qnh,
    humedad = hr,
    viento_vel_m_s = viento_vel_m_s,
    visibilidad = NA_real_,
    estacion = "AEROPUERTO"
  )

# Bind all stations and convert to long format (2024 only)
meteo_all <- bind_rows(
  oya518_sel, oya780_sel, oya862_sel, oya1069_sel,
  oya1128_sel, oya1193_sel, oya1211_sel, oya1354_sel,
  airport_sel
)

meteo_long <- meteo_all %>%
  pivot_longer(cols = -c(datetime, estacion), names_to = "variable", values_to = "valor") %>%
  filter(year(datetime) == 2024)

meteo_long <- bind_rows(meteo_long, goes)
```

# Cleaning rules

The following values are treated as missing (logger artefacts):

- air temperature equal to 0 °C,
- air pressure equal to 0 hPa,
- wind speed equal to 0 m/s,
- visibility below 0.4 m.

```{r}
#| label: cleaning

meteo_long <- meteo_long %>%
  mutate(
    valor = case_when(
      variable == "temperatura"    & valor == 0  ~ NA_real_,
      variable == "presion"        & valor == 0  ~ NA_real_,
      variable == "viento_vel_m_s" & valor == 0  ~ NA_real_,
      variable == "visibilidad"    & valor < 0.4 ~ NA_real_,
      TRUE ~ valor
    )
  )
```

# Fog and dew water

Fog (`niebla`) and dew (`rocio`) water are recorded in mm and/or mL depending on the station and period. Both records are merged into a single variable per phenomenon (`niebla_final`, `rocio_final`) taking the mm value when available and the mL value otherwise, without unit conversion.

```{r}
#| label: fog-dew-water

meteo_check <- meteo_all %>%
  mutate(
    fog_both    = !is.na(niebla_mm) & !is.na(niebla_ml),
    fog_only_mm = !is.na(niebla_mm) &  is.na(niebla_ml),
    fog_only_ml =  is.na(niebla_mm) & !is.na(niebla_ml),
    dew_both    = !is.na(rocio_mm)  & !is.na(rocio_ml),
    dew_only_mm = !is.na(rocio_mm)  &  is.na(rocio_ml),
    dew_only_ml =  is.na(rocio_mm)  & !is.na(rocio_ml)
  )

meteo_check %>%
  group_by(estacion) %>%
  summarise(rows = n(),
            fog_mm_and_ml = sum(fog_both), fog_only_mm = sum(fog_only_mm), fog_only_ml = sum(fog_only_ml),
            dew_mm_and_ml = sum(dew_both), dew_only_mm = sum(dew_only_mm), dew_only_ml = sum(dew_only_ml)) %>%
  kable(caption = "Number of rows with fog and dew water in mm and/or mL per station")

water_long <- meteo_all %>%
  mutate(
    niebla_final = coalesce(niebla_mm, niebla_ml),
    rocio_final  = coalesce(rocio_mm, rocio_ml)
  ) %>%
  select(datetime, estacion, niebla_final, rocio_final) %>%
  pivot_longer(cols = c(niebla_final, rocio_final), names_to = "variable", values_to = "valor") %>%
  filter(!is.na(valor))

meteo_long <- meteo_long %>%
  filter(!variable %in% c("niebla_mm", "niebla_ml", "rocio_mm", "rocio_ml")) %>%
  bind_rows(water_long)
```

# Interactive check of each variable

```{r}
#| label: plots

plot_variable <- function(data, var_name, mode_type = "lines") {
  data %>%
    filter(variable == var_name,
           datetime >= as.POSIXct("2024-01-01", tz = "UTC"),
           datetime <  as.POSIXct("2025-01-01", tz = "UTC")) %>%
    plot_ly(x = ~datetime, y = ~valor, color = ~estacion,
            type = "scatter", mode = mode_type) %>%
    layout(title = var_name, xaxis = list(title = "Date"), yaxis = list(title = var_name))
}

variables_plot <- c("temperatura", "presion", "humedad", "viento_vel_m_s", "viento_dir",
                    "pp_mm", "radiacion_w_m2", "temp_sup_c", "visibilidad",
                    "niebla_final", "rocio_final")

htmltools::tagList(lapply(variables_plot, function(v) plot_variable(meteo_long, v)))
plot_variable(meteo_long, "FLC", mode_type = "markers")
```

# Data availability (Table 1)

```{r}
#| label: availability

expected_2024 <- 366 * 24 * 6  # 2024 is a leap year, 10-min records = 52704

availability <- meteo_long %>%
  filter(year(datetime) == 2024) %>%
  group_by(variable, estacion) %>%
  summarise(records = sum(!is.na(valor)),
            availability_pct = round(100 * records / expected_2024, 1),
            .groups = "drop") %>%
  select(variable, estacion, availability_pct) %>%
  pivot_wider(names_from = estacion, values_from = availability_pct) %>%
  arrange(variable) %>%
  select(any_of(c("variable", "GOES", "AEROPUERTO", "OYA_518", "OYA_780", "OYA_862",
                  "OYA_1069", "OYA_1128", "OYA_1193", "OYA_1211", "OYA_1354")))

availability %>%
  kable(format = "html", caption = "Data availability (%) in 2024 at 10-min resolution") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
#| label: save

meteo_long %>%
  filter(year(datetime) == 2024) %>%
  pivot_wider(names_from = variable, values_from = valor) %>%
  write_csv(file.path(dir_data, "derived/stations_10min_2024.csv"))
```
