---
title: "03 Classification of fog and dew occurrences at OYA1211 (visibility + GOES-16)"
author: "Francisca Muñoz Narbona"
format:
  html:
    toc: true
    self-contained: true
editor: visual
editor_options:
  chunk_output_type: console
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
    dpi: 300
---

Step 3 of the pipeline. Classifies every 10-min record of the main station OYA1211 during 2024 as fog (`NIEBLA`), dew (`ROCIO`) or no occurrence (`SIN_EVENTO`), following the hierarchical criterion of the manuscript (Eq. 1):

1. If neither collector (SFC or SDC) registers water, the record is `SIN_EVENTO`.
2. If visibility is available: fog when visibility < 1000 m, dew when visibility >= 1000 m.
3. If visibility is unavailable: fog when GOES-16 detects fog and low clouds (FLC = 1), dew otherwise.

The automatic classification is then reviewed day by day against the visibility curve and the GOES signal, and corrected in the time windows listed in `manual_corrections`. The output is `data/derived/oya1211_classification_2024.csv`.

```{r}
#| label: automatic-classification

library(tidyverse)
library(lubridate)
library(knitr)

dir_data <- "../data"

data <- read_csv(file.path(dir_data, "derived/stations_10min_2024.csv"))

oya1211 <- data %>%
  filter(estacion == "OYA_1211")

goes <- data %>%
  filter(estacion == "GOES") %>%
  mutate(
    datetime = as.POSIXct(datetime, tz = "UTC"),
    ts_10    = floor_date(datetime, "10 minutes"),
    fog_sat  = if_else(!is.na(FLC) & FLC == 1, 1L, 0L)
  ) %>%
  select(ts_10, fog_sat)

oya1211_class_auto <- oya1211 %>%
  mutate(
    datetime = as.POSIXct(datetime, tz = "UTC"),
    ts_10    = floor_date(datetime, "10 minutes"),
    fecha    = as.Date(ts_10),

    vis_m = visibilidad,

    # instrumental collection: 1 if the collector registered water
    fog_cap = if_else(!is.na(niebla_final) & niebla_final > 0, 1L, 0L),
    dew_cap = if_else(!is.na(rocio_final)  & rocio_final  > 0, 1L, 0L),
    any_cap = if_else(fog_cap == 1L | dew_cap == 1L, 1L, 0L)
  ) %>%
  left_join(goes, by = "ts_10") %>%
  mutate(
    clasif_evento = case_when(
      any_cap == 0L ~ "SIN_EVENTO",

      !is.na(vis_m) & vis_m < 1000  ~ "NIEBLA",
      !is.na(vis_m) & vis_m >= 1000 ~ "ROCIO",

      is.na(vis_m) & fog_sat == 1L ~ "NIEBLA",
      is.na(vis_m) & fog_sat == 0L ~ "ROCIO",

      TRUE ~ NA_character_
    )
  ) %>%
  filter(year(fecha) == 2024)
```

# Visual review (half-month panels)

Visibility (blue line; red dashed line, 1000 m threshold), the classification band and the GOES FLC band for every day of 2024, used for the manual review.

```{r}
#| label: review-plot-function

plot_half_month <- function(df, month_int, half_month = 1, facet_ncol = 4) {

  dsub <- df %>%
    filter(month(fecha) == month_int) %>%
    filter(if (half_month == 1) day(fecha) <= 15 else day(fecha) > 15)

  if (nrow(dsub) == 0) return(NULL)

  band_h <- 500
  y_cls  <- -500
  y_goes <- -1200

  dsub <- dsub %>%
    mutate(class_fac = factor(clasif_evento, levels = c("NIEBLA", "ROCIO", "SIN_EVENTO")))

  ggplot(dsub, aes(x = ts_10)) +
    geom_tile(aes(y = y_goes, alpha = fog_sat), fill = "grey30", height = band_h, width = 600) +
    scale_alpha_continuous(range = c(0.15, 0.9), na.value = 0, guide = "none") +
    geom_tile(aes(y = y_cls, fill = class_fac), height = band_h, width = 600) +
    scale_fill_manual(values = c("NIEBLA" = "#6baed6", "ROCIO" = "#c97a4a", "SIN_EVENTO" = "#bdbdbd"),
                      labels = c("NIEBLA" = "Fog", "ROCIO" = "Dew", "SIN_EVENTO" = "No occurrence"),
                      name = "Classification") +
    geom_line(aes(y = vis_m, group = fecha), color = "#08519c", linewidth = 0.35, na.rm = TRUE) +
    geom_hline(yintercept = 1000, linetype = "dashed", color = "red", linewidth = 0.35) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    scale_y_continuous(limits = c(-1700, 5000),
                       breaks = c(-1200, -500, 0, 1000, 2000, 3000, 4000, 5000),
                       labels = c("GOES", "Classif.", "0", "1000", "2000", "3000", "4000", "5000"),
                       expand = expansion(mult = c(0.01, 0.03))) +
    scale_x_datetime(date_breaks = "2 hours", date_labels = "%H:%M") +
    facet_wrap(~ fecha, ncol = facet_ncol, scales = "free_x") +
    labs(x = "Time (UTC)", y = "Visibility (m)",
         title = paste0("Visibility and classification, ", year(min(dsub$fecha)), "-",
                        sprintf("%02d", month_int), " (half ", half_month, ")")) +
    theme_bw(base_size = 8) +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold"),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
          axis.text.y = element_text(face = "bold"),
          plot.margin = margin(6, 6, 6, 42))
}
```

```{r}
#| label: review-automatic
#| fig-width: 9
#| fig-height: 9

for (m in 1:12) {
  print(plot_half_month(oya1211_class_auto, m, half_month = 1))
  print(plot_half_month(oya1211_class_auto, m, half_month = 2))
}
```

# Manual corrections

Time windows (UTC) reclassified after the visual review. A correction only applies to records where at least one collector registered water.

```{r}
#| label: manual-corrections

manual_corrections <- tribble(
  ~start, ~end, ~occurrence,

  ymd_hms("2024-01-02 02:00:00", tz = "UTC"),
  ymd_hms("2024-01-02 03:30:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-01-02 10:00:00", tz = "UTC"),
  ymd_hms("2024-01-02 12:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-01-08 10:00:00", tz = "UTC"),
  ymd_hms("2024-01-08 12:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-01-11 03:00:00", tz = "UTC"),
  ymd_hms("2024-01-11 06:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-01-15 09:00:00", tz = "UTC"),
  ymd_hms("2024-01-15 12:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-02-03 02:00:00", tz = "UTC"),
  ymd_hms("2024-02-03 05:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-01-28 12:00:00", tz = "UTC"),
  ymd_hms("2024-01-28 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-01-31 09:00:00", tz = "UTC"),
  ymd_hms("2024-01-31 22:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-02-03 07:00:00", tz = "UTC"),
  ymd_hms("2024-02-03 11:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-02-04 02:00:00", tz = "UTC"),
  ymd_hms("2024-02-04 12:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-02-12 06:00:00", tz = "UTC"),
  ymd_hms("2024-02-12 10:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-02-17 05:00:00", tz = "UTC"),
  ymd_hms("2024-02-17 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-02-18 11:00:00", tz = "UTC"),
  ymd_hms("2024-02-18 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-02-19 04:00:00", tz = "UTC"),
  ymd_hms("2024-02-19 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-02-28 10:00:00", tz = "UTC"),
  ymd_hms("2024-02-28 13:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-03-07 08:00:00", tz = "UTC"),
  ymd_hms("2024-03-07 14:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-03-09 01:00:00", tz = "UTC"),
  ymd_hms("2024-03-09 13:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-03-09 08:00:00", tz = "UTC"),
  ymd_hms("2024-03-09 15:00:00", tz = "UTC"),
  "NIEBLA",

  # uncertain day(s): 10 March

  ymd_hms("2024-03-13 04:00:00", tz = "UTC"),
  ymd_hms("2024-03-13 16:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-03-16 06:00:00", tz = "UTC"),
  ymd_hms("2024-03-16 08:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-03-24 14:00:00", tz = "UTC"),
  ymd_hms("2024-03-24 16:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-03-25 02:00:00", tz = "UTC"),
  ymd_hms("2024-03-25 15:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-03-26 07:00:00", tz = "UTC"),
  ymd_hms("2024-03-26 09:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-03-26 11:00:00", tz = "UTC"),
  ymd_hms("2024-03-26 14:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-02 08:00:00", tz = "UTC"),
  ymd_hms("2024-04-02 14:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-07 00:00:00", tz = "UTC"),
  ymd_hms("2024-04-07 04:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-04-07 06:00:00", tz = "UTC"),
  ymd_hms("2024-04-07 15:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-08 00:00:00", tz = "UTC"),
  ymd_hms("2024-04-08 04:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-04-09 09:00:00", tz = "UTC"),
  ymd_hms("2024-04-09 11:00:00", tz = "UTC"),
  "ROCIO",

  # NOTE: the end time below ("2024-04-1") parses as 1 April, before the start time,
  # so this window has no effect; it is kept as is to reproduce the classification
  # used in the manuscript (probably meant 2024-04-11 03:00).
  ymd_hms("2024-04-11 00:00:00", tz = "UTC"),
  ymd_hms("2024-04-1 03:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-13 08:00:00", tz = "UTC"),
  ymd_hms("2024-04-13 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-15 06:00:00", tz = "UTC"),
  ymd_hms("2024-04-15 07:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-22 14:00:00", tz = "UTC"),
  ymd_hms("2024-04-22 23:50:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-23 04:00:00", tz = "UTC"),
  ymd_hms("2024-04-23 06:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 26 April

  ymd_hms("2024-04-27 00:00:00", tz = "UTC"),
  ymd_hms("2024-04-27 14:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-04-28 04:00:00", tz = "UTC"),
  ymd_hms("2024-04-28 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-05-04 00:00:00", tz = "UTC"),
  ymd_hms("2024-05-04 05:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-05-05 00:00:00", tz = "UTC"),
  ymd_hms("2024-05-05 03:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-05-07 03:00:00", tz = "UTC"),
  ymd_hms("2024-05-07 12:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-05-08 00:00:00", tz = "UTC"),
  ymd_hms("2024-05-08 01:30:00", tz = "UTC"),
  "NIEBLA",

  # uncertain day(s): 10 May

  ymd_hms("2024-05-11 00:00:00", tz = "UTC"),
  ymd_hms("2024-05-11 23:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 12 May

  ymd_hms("2024-05-13 00:00:00", tz = "UTC"),
  ymd_hms("2024-05-13 16:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-05-16 00:00:00", tz = "UTC"),
  ymd_hms("2024-05-16 12:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 17 May

  ymd_hms("2024-05-18 00:00:00", tz = "UTC"),
  ymd_hms("2024-05-19 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-06-09 00:00:00", tz = "UTC"),
  ymd_hms("2024-06-09 23:50:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-07-08 00:00:00", tz = "UTC"),
  ymd_hms("2024-07-08 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-07-11 00:00:00", tz = "UTC"),
  ymd_hms("2024-07-11 05:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-07-18 00:00:00", tz = "UTC"),
  ymd_hms("2024-07-18 23:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 06 August

  ymd_hms("2024-08-22 06:00:00", tz = "UTC"),
  ymd_hms("2024-08-22 07:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 21 August

  ymd_hms("2024-08-30 00:00:00", tz = "UTC"),
  ymd_hms("2024-08-30 08:30:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-08-31 00:00:00", tz = "UTC"),
  ymd_hms("2024-08-31 12:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 06 September

  ymd_hms("2024-09-13 00:00:00", tz = "UTC"),
  ymd_hms("2024-09-13 08:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-10-08 00:00:00", tz = "UTC"),
  ymd_hms("2024-10-08 14:00:00", tz = "UTC"),
  "ROCIO",

  # NOTE (to be reviewed): during this window visibility is ~500 m and the fog
  # collector registers ~4.7 L, and an earlier version of this list classified it
  # as NIEBLA. It is kept as ROCIO to reproduce the classification used in the
  # manuscript.
  ymd_hms("2024-10-08 16:00:00", tz = "UTC"),
  ymd_hms("2024-10-08 23:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 09 October

  ymd_hms("2024-10-10 00:00:00", tz = "UTC"),
  ymd_hms("2024-10-10 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-10-19 00:00:00", tz = "UTC"),
  ymd_hms("2024-10-19 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-10-21 00:00:00", tz = "UTC"),
  ymd_hms("2024-10-21 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-10-31 00:00:00", tz = "UTC"),
  ymd_hms("2024-10-31 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-11-01 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-01 08:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-11-01 12:00:00", tz = "UTC"),
  ymd_hms("2024-11-01 23:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-11-02 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-02 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-11-07 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-07 23:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-11-08 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-08 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-11-09 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-09 05:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-11-09 12:00:00", tz = "UTC"),
  ymd_hms("2024-11-09 23:00:00", tz = "UTC"),
  "NIEBLA",

  # uncertain day(s): 10 October

  ymd_hms("2024-11-13 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-13 23:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-11-15 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-15 23:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 18 to 22 November

  ymd_hms("2024-11-25 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-25 05:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-11-27 10:00:00", tz = "UTC"),
  ymd_hms("2024-11-27 23:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-11-29 00:00:00", tz = "UTC"),
  ymd_hms("2024-11-29 23:00:00", tz = "UTC"),
  "ROCIO",

  ymd_hms("2024-12-03 09:00:00", tz = "UTC"),
  ymd_hms("2024-12-03 23:00:00", tz = "UTC"),
  "NIEBLA",

  # uncertain day(s): 09 December

  ymd_hms("2024-12-19 00:00:00", tz = "UTC"),
  ymd_hms("2024-12-19 13:00:00", tz = "UTC"),
  "NIEBLA",

  ymd_hms("2024-12-23 00:00:00", tz = "UTC"),
  ymd_hms("2024-12-23 23:00:00", tz = "UTC"),
  "ROCIO",

  # uncertain day(s): 27 December

  ymd_hms("2024-12-28 02:00:00", tz = "UTC"),
  ymd_hms("2024-12-28 06:00:00", tz = "UTC"),
  "ROCIO"

)

oya1211_class <- oya1211_class_auto

for (i in seq_len(nrow(manual_corrections))) {
  oya1211_class <- oya1211_class %>%
    mutate(
      clasif_evento = if_else(
        datetime >= manual_corrections$start[i] &
          datetime <= manual_corrections$end[i] &
          any_cap == 1,
        manual_corrections$occurrence[i],
        clasif_evento
      )
    )
}
```

```{r}
#| label: review-corrected
#| fig-width: 9
#| fig-height: 9

for (m in 1:12) {
  print(plot_half_month(oya1211_class, m, half_month = 1))
  print(plot_half_month(oya1211_class, m, half_month = 2))
}
```

# Summary and output

```{r}
#| label: summary-save

# share of the 10-min records of 2024 in each category, before and after the classification
bind_rows(
  oya1211_class_auto %>%
    mutate(type = case_when(fog_cap == 1 & dew_cap == 1 ~ "Fog + dew",
                            fog_cap == 1 ~ "Fog", dew_cap == 1 ~ "Dew",
                            TRUE ~ "No occurrence")) %>%
    count(type) %>% mutate(scheme = "Instrumental collection"),
  oya1211_class %>%
    mutate(type = recode(clasif_evento, "NIEBLA" = "Fog", "ROCIO" = "Dew",
                         "SIN_EVENTO" = "No occurrence")) %>%
    count(type) %>% mutate(scheme = "Visibility + GOES classification")
) %>%
  group_by(scheme) %>%
  mutate(pct = round(100 * n / sum(n), 2)) %>%
  ungroup() %>%
  kable()

write_csv(oya1211_class, file.path(dir_data, "derived/oya1211_classification_2024.csv"))
```
