---
title: "05 Temporal tendencies of q, θ and U at OYA1211 (Section 3.3)"
author: "Francisca Muñoz Narbona"
format:
  html:
    toc: true
    self-contained: true
editor: visual
editor_options:
  chunk_output_type: console
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
    dpi: 300
---

Step 5 of the pipeline. Uses `data/derived/stations_theta_q_2024.csv` (step 4) at the main station OYA1211 to describe the temporal evolution of specific humidity (q), potential temperature (θ) and wind speed (U) under fog, dew and no occurrence:

1. **Change after each record** (Fig. 7 of the manuscript): for every 10-min record, the change of q, θ and U after fixed lags of 10 to 180 min, averaged by occurrence type. Note that the lags are computed from every record of each type, not only from the first record (onset) of each occurrence.
2. **Diurnal cycle of the temporal rates** (Fig. 8): forward finite differences ∂φ/∂t computed with the first valid value between 1 and 2 h ahead, averaged by time of day (10-min slots) and occurrence type.
3. **Distribution of the temporal rates** (Fig. 9): same rates with the first valid value between 1 and 3 h ahead, after removing outliers (1.5 × IQR) per variable and occurrence type.

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(zoo)

dir_data <- "../data"
dir_fig  <- "../figures"

theta_q <- read_csv(file.path(dir_data, "derived/stations_theta_q_2024.csv"), show_col_types = FALSE)

data_1211 <- theta_q %>%
  filter(estacion == "OYA_1211") %>%
  arrange(datetime)

col_occ <- c("Fog" = "#4292c6", "Dew" = "#c97a4a", "No occurrence" = "#7f7f7f")

label_occurrence <- function(x) {
  factor(recode(x, "NIEBLA" = "Fog", "ROCIO" = "Dew", "SIN_EVENTO" = "No occurrence"),
         levels = c("Fog", "Dew", "No occurrence"))
}

theme_fig <- theme_classic(base_size = 10) +
  theme(
    legend.title     = element_blank(),
    legend.position  = "bottom",
    axis.line        = element_line(color = "#c3c2b7"),
    axis.ticks       = element_line(color = "#c3c2b7"),
    panel.grid.major.y = element_line(color = "#ecebe6", linewidth = 0.3),
    strip.background = element_rect(fill = "#f9f9f7", color = NA),
    strip.text       = element_text(face = "bold")
  )
```

# 1. Change after each record (lags of 10 to 180 min)

```{r}
#| label: fig-change-after-record
#| fig-width: 7
#| fig-height: 8

data_lag <- data_1211

for (i in 1:18) {
  data_lag[[paste0("q_",  i * 10)]] <- dplyr::lead(data_lag$q_kgkg, i)
  data_lag[[paste0("th_", i * 10)]] <- dplyr::lead(data_lag$theta_K, i)
  data_lag[[paste0("ws_", i * 10)]] <- dplyr::lead(data_lag$viento_vel_m_s, i)
}

for (i in seq(10, 180, 10)) {
  data_lag[[paste0("dq_", i)]]  <- (data_lag[[paste0("q_", i)]] - data_lag$q_kgkg) * 1000  # g/kg
  data_lag[[paste0("dth_", i)]] <- data_lag[[paste0("th_", i)]] - data_lag$theta_K       # K
  data_lag[[paste0("dws_", i)]] <- data_lag[[paste0("ws_", i)]] - data_lag$viento_vel_m_s # m/s
}

change_mean <- data_lag %>%
  select(clasif_evento, starts_with("dq_"), starts_with("dth_"), starts_with("dws_")) %>%
  pivot_longer(cols = -clasif_evento, names_to = "var_lag", values_to = "delta") %>%
  mutate(
    variable = case_when(grepl("^dq_", var_lag)  ~ "q",
                         grepl("^dth_", var_lag) ~ "theta",
                         grepl("^dws_", var_lag) ~ "wind"),
    lag_min = as.numeric(gsub("[a-z_]+", "", var_lag))
  ) %>%
  group_by(variable, clasif_evento, lag_min) %>%
  summarise(mean_delta = mean(delta, na.rm = TRUE),
            sd_delta   = sd(delta, na.rm = TRUE),
            .groups = "drop") %>%
  filter(!is.na(clasif_evento)) %>%
  mutate(
    variable = factor(variable, levels = c("q", "theta", "wind"),
                      labels = c("Specific humidity, Δq (g kg⁻¹)",
                                 "Potential temperature, Δθ (K)",
                                 "Wind speed, ΔU (m s⁻¹)")),
    occurrence = label_occurrence(clasif_evento)
  )

fig_change <- ggplot(change_mean, aes(x = lag_min, y = mean_delta, color = occurrence)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = col_occ) +
  scale_x_continuous(breaks = seq(0, 180, 30)) +
  labs(x = "Minutes after the record", y = "Mean change") +
  theme_fig

fig_change

ggsave(file.path(dir_fig, "fig07_change_after_record.png"),
       fig_change, width = 7, height = 8, dpi = 300, bg = "white")

# values used in the text (180 min)
change_mean %>% filter(lag_min == 180) %>% select(variable, occurrence, mean_delta)
```

# 2. Diurnal cycle of the temporal rates (1 to 2 h)

```{r}
#| label: rates-function

# Forward finite difference: first valid value between lag_min and lag_max
# 10-min steps ahead (6 = 1 h), divided by the elapsed time in hours
forward_rate <- function(i, var_vec, time_vec, lag_min = 6, lag_max = 12) {
  for (lag in lag_min:lag_max) {
    j <- i + lag
    if (j <= length(var_vec) && !is.na(var_vec[j])) {
      dt_hours <- as.numeric(difftime(time_vec[j], time_vec[i], units = "hours"))
      return((var_vec[j] - var_vec[i]) / dt_hours)
    }
  }
  NA_real_
}

compute_rates <- function(df, lag_max) {
  time_vec <- df$datetime
  df %>%
    mutate(
      dq_dt  = sapply(seq_along(q_kgkg), forward_rate, var_vec = q_kgkg,
                      time_vec = time_vec, lag_max = lag_max) * 1000,          # g/kg h-1
      dth_dt = sapply(seq_along(theta_K), forward_rate, var_vec = theta_K,
                      time_vec = time_vec, lag_max = lag_max),                 # K h-1
      dws_dt = sapply(seq_along(viento_vel_m_s), forward_rate, var_vec = viento_vel_m_s,
                      time_vec = time_vec, lag_max = lag_max)                  # m/s h-1
    )
}

rate_labels <- c("dq_dt"  = "Specific humidity, ∂q/∂t (g kg⁻¹ h⁻¹)",
                 "dth_dt" = "Potential temperature, ∂θ/∂t (K h⁻¹)",
                 "dws_dt" = "Wind speed, ∂U/∂t (m s⁻¹ h⁻¹)")
```

```{r}
#| label: fig-diurnal-rates
#| fig-width: 8
#| fig-height: 8

rates_2h <- compute_rates(data_1211, lag_max = 12) %>%
  mutate(slot_10min = (hour(datetime) * 60 + minute(datetime)) / 10)

diurnal <- rates_2h %>%
  select(clasif_evento, slot_10min, dq_dt, dth_dt, dws_dt) %>%
  pivot_longer(cols = c(dq_dt, dth_dt, dws_dt), names_to = "variable", values_to = "rate") %>%
  group_by(variable, clasif_evento, slot_10min) %>%
  summarise(mean_rate = mean(rate, na.rm = TRUE),
            sd_rate   = sd(rate, na.rm = TRUE),
            .groups = "drop") %>%
  filter(!is.na(clasif_evento)) %>%
  arrange(variable, clasif_evento, slot_10min) %>%
  group_by(variable, clasif_evento) %>%
  mutate(mean_rate = na.approx(mean_rate, na.rm = FALSE),   # fill isolated empty slots
         sd_rate   = na.approx(sd_rate, na.rm = FALSE)) %>%
  ungroup() %>%
  mutate(hour = slot_10min / 6,
         variable = factor(variable, levels = names(rate_labels), labels = rate_labels),
         occurrence = label_occurrence(clasif_evento))

fig_diurnal <- ggplot(diurnal, aes(x = hour, y = mean_rate, color = occurrence, fill = occurrence)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_ribbon(aes(ymin = mean_rate - sd_rate, ymax = mean_rate + sd_rate), alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = col_occ) +
  scale_fill_manual(values = col_occ) +
  scale_x_continuous(
    breaks = seq(0, 24, 3), limits = c(0, 24), name = "Hour of day (UTC)",
    sec.axis = sec_axis(~ ., breaks = seq(0, 24, 3), labels = (seq(0, 24, 3) - 4) %% 24,
                        name = "Chile local time (UTC-4)")
  ) +
  labs(y = "Mean rate (± 1 SD)") +
  theme_fig

fig_diurnal

ggsave(file.path(dir_fig, "fig08_diurnal_rates.png"),
       fig_diurnal, width = 8, height = 8, dpi = 300, bg = "white")
```

# 3. Distribution of the temporal rates (1 to 3 h)

```{r}
#| label: fig-rate-distributions
#| fig-width: 10
#| fig-height: 5

rates_3h <- compute_rates(data_1211, lag_max = 18)

rates_clean <- rates_3h %>%
  select(clasif_evento, dq_dt, dth_dt, dws_dt) %>%
  pivot_longer(cols = -clasif_evento, names_to = "variable", values_to = "rate") %>%
  filter(!is.na(clasif_evento)) %>%
  group_by(variable, clasif_evento) %>%
  mutate(Q1 = quantile(rate, 0.25, na.rm = TRUE),
         Q3 = quantile(rate, 0.75, na.rm = TRUE),
         IQR = Q3 - Q1) %>%
  filter(rate >= Q1 - 1.5 * IQR, rate <= Q3 + 1.5 * IQR) %>%
  ungroup() %>%
  mutate(variable = factor(variable, levels = names(rate_labels), labels = rate_labels),
         occurrence = label_occurrence(clasif_evento))

fig_distributions <- ggplot(rates_clean, aes(x = occurrence, y = rate, fill = occurrence)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_violin(alpha = 0.35, color = NA, trim = FALSE) +
  geom_boxplot(width = 0.18, alpha = 0.9, outlier.shape = NA, color = "grey20") +
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = col_occ) +
  labs(x = NULL, y = "Temporal rate") +
  theme_fig

fig_distributions

ggsave(file.path(dir_fig, "fig09_rate_distributions.png"),
       fig_distributions, width = 10, height = 5, dpi = 300, bg = "white")

# medians per occurrence type
rates_clean %>% group_by(variable, occurrence) %>%
  summarise(median = median(rate), n = n(), .groups = "drop")
```
