---
title: "Figures of Section 3.1: monthly and hourly distribution of fog and dew at OYA1211"
author: "Francisca Muñoz Narbona"
format:
  html:
    toc: true
    self-contained: true
editor: visual
editor_options:
  chunk_output_type: console
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
    dpi: 300
---

Figure 4 of the manuscript. Four panels: monthly fog, monthly dew, hourly fog and hourly dew at OYA1211. In each panel the bars are the mean accumulated water (left axis, mL; mean of the 10-min values > 0; instrumental in grey and classified in the colour of the phenomenon) and the lines are the frequency (right axis, %; instrumental dashed with open points, classified solid with filled points). Each panel has its own water scale; the frequency scale is always 0 to 100%.

- Instrumental: according to the collector that registered water (fog collector, SFC; dew collector, SDC).
- Classified: visibility + GOES-16 classification with manual corrections (step 3).
- Frequency: % of the 10-min records with atmospheric water in each month or hour. Classified fog + dew add up to 100%; instrumental fog and dew both include the records in which both collectors registered water, so they can add up to more than 100% (this is the instrumental overlap).

Input: `data/derived/oya1211_classification_2024.csv` (step 3). Output: `figures/fig04_seasonal_diurnal_distribution.png`.

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)

dir_data <- "../data"
dir_fig  <- "../figures"
```

# Data: instrumental vs classified

```{r}
#| label: data

oya <- read_csv(
  file.path(dir_data, "derived/oya1211_classification_2024.csv"),
  col_types = cols(datetime = col_character()),
  show_col_types = FALSE
) %>%
  mutate(datetime = ymd_hms(datetime, tz = "UTC")) %>%
  filter(!is.na(niebla_final), !is.na(rocio_final))

months_en <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

oya <- oya %>%
  mutate(
    water_total = niebla_final + rocio_final,
    collection  = water_total > 0,
    month       = factor(months_en[month(datetime)], levels = months_en),
    hour_num    = hour(datetime),
    # classified scheme: final classification of step 3
    type_final  = recode(clasif_evento, "NIEBLA" = "Fog", "ROCIO" = "Dew",
                         "SIN_EVENTO" = "No occurrence")
  )

# Mean of the values > 0 per group (month or hour)
mean_by <- function(df, group, scheme_label) {
  df %>%
    filter(value > 0) %>%
    group_by({{ group }}) %>%
    summarise(mean = mean(value), n_obs = n(), .groups = "drop") %>%
    mutate(scheme = scheme_label)
}

# Mean accumulated water per group: instrumental (water of the collector of
# each phenomenon) and classified (total water of the records of each class)
water_by <- function(group) {
  bind_rows(
    mean_by(oya %>% filter(niebla_final > 0) %>% transmute({{ group }}, value = niebla_final),
            {{ group }}, "Instrumental") %>% mutate(phenomenon = "Fog"),
    mean_by(oya %>% filter(type_final == "Fog") %>% transmute({{ group }}, value = water_total),
            {{ group }}, "Classified") %>% mutate(phenomenon = "Fog"),
    mean_by(oya %>% filter(rocio_final > 0) %>% transmute({{ group }}, value = rocio_final),
            {{ group }}, "Instrumental") %>% mutate(phenomenon = "Dew"),
    mean_by(oya %>% filter(type_final == "Dew") %>% transmute({{ group }}, value = water_total),
            {{ group }}, "Classified") %>% mutate(phenomenon = "Dew")
  ) %>%
    transmute({{ group }}, value = mean, phenomenon, criterion = scheme)
}

# Frequency (%) per group over all the records with atmospheric water
freq_by <- function(group) {
  base  <- oya %>% filter(collection)
  total <- base %>% count({{ group }}, name = "total")
  bind_rows(
    base %>% filter(niebla_final > 0) %>% count({{ group }}, name = "n") %>%
      mutate(phenomenon = "Fog", criterion = "Instrumental"),
    base %>% filter(rocio_final > 0) %>% count({{ group }}, name = "n") %>%
      mutate(phenomenon = "Dew", criterion = "Instrumental"),
    base %>% filter(type_final == "Fog") %>% count({{ group }}, name = "n") %>%
      mutate(phenomenon = "Fog", criterion = "Classified"),
    base %>% filter(type_final == "Dew") %>% count({{ group }}, name = "n") %>%
      mutate(phenomenon = "Dew", criterion = "Classified")
  ) %>%
    left_join(total, by = join_by({{ group }})) %>%
    mutate(value = n / total * 100)
}

levels_water <- c("Water, instrumental", "Water, classified (fog)", "Water, classified (dew)")
levels_freq  <- c("Frequency, instrumental", "Frequency, classified (fog)", "Frequency, classified (dew)")

add_key <- function(df, kind) {
  df %>% mutate(
    phenomenon = factor(phenomenon, levels = c("Fog", "Dew")),
    key = case_when(
      criterion == "Instrumental" ~ paste0(kind, ", instrumental"),
      phenomenon == "Fog"         ~ paste0(kind, ", classified (fog)"),
      TRUE                        ~ paste0(kind, ", classified (dew)")
    ),
    key = factor(key, levels = if (kind == "Water") levels_water else levels_freq)
  )
}

water_month <- water_by(month)    %>% add_key("Water")
water_hour  <- water_by(hour_num) %>% add_key("Water")
freq_month  <- freq_by(month)     %>% add_key("Frequency")
freq_hour   <- freq_by(hour_num)  %>% add_key("Frequency")
```

# Figure

```{r}
#| label: figure-settings

col_water <- c("Water, instrumental"     = "#bdbdbd",
               "Water, classified (fog)" = "#4292c6",
               "Water, classified (dew)" = "#c97a4a")
col_freq  <- c("Frequency, instrumental"     = "#252525",
               "Frequency, classified (fog)" = "#08306b",
               "Frequency, classified (dew)" = "#7f2704")
lt_freq   <- c("Frequency, instrumental"     = "22",
               "Frequency, classified (fog)" = "solid",
               "Frequency, classified (dew)" = "solid")
sh_freq   <- c("Frequency, instrumental"     = 1,
               "Frequency, classified (fog)" = 16,
               "Frequency, classified (dew)" = 16)

theme_comb <- theme_classic(base_size = 10) +
  theme(
    legend.title    = element_blank(),
    legend.position = "bottom",
    axis.line       = element_line(color = "#c3c2b7"),
    axis.ticks      = element_line(color = "#c3c2b7"),
    panel.grid.major.y = element_line(color = "#ecebe6", linewidth = 0.3),
    plot.title      = element_text(face = "bold", size = 10)
  )

panel <- function(water, freq, x, phen, title, hourly = FALSE) {
  a <- water %>% filter(phenomenon == phen)
  f <- freq  %>% filter(phenomenon == phen)
  k <- max(a$value, na.rm = TRUE) * 1.05 / 100   # 100% frequency = top of the water axis

  p <- ggplot() +
    geom_col(data = a, aes(x = {{ x }}, y = value, fill = key),
             position = position_dodge(width = 0.8, preserve = "single"),
             width = 0.75, na.rm = TRUE) +
    geom_line(data = f, aes(x = {{ x }}, y = value * k, color = key,
                            linetype = key, group = key),
              linewidth = 0.5, na.rm = TRUE) +
    geom_point(data = f, aes(x = {{ x }}, y = value * k, color = key, shape = key),
               size = 1.5, stroke = 0.6, na.rm = TRUE) +
    scale_fill_manual(values = col_water, drop = FALSE) +
    scale_color_manual(values = col_freq, drop = FALSE) +
    scale_linetype_manual(values = lt_freq, drop = FALSE) +
    scale_shape_manual(values = sh_freq, drop = FALSE) +
    scale_y_continuous(
      name = "Mean accumulated water (mL)",
      limits = c(0, 100 * k),
      expand = expansion(mult = c(0, 0.02)),
      sec.axis = sec_axis(transform = ~ . / k, name = "Frequency (%)",
                          breaks = seq(0, 100, 25))
    ) +
    labs(title = title, x = if (hourly) "Hour of day (UTC)" else NULL) +
    theme_comb +
    theme(legend.box = "vertical", legend.spacing.y = unit(0, "pt"))

  if (hourly) {
    p <- p + scale_x_continuous(
      breaks = 0:23, labels = sprintf("%02d", 0:23),
      expand = expansion(add = 0.6),
      sec.axis = sec_axis(transform = ~ ., breaks = 0:23,
                          labels = sprintf("%02d", (0:23 - 3) %% 24),
                          name = "Chile local time (approx., UTC-3)")
    ) +
      theme(axis.text.x = element_text(size = 6.5),
            axis.text.x.top = element_text(size = 6.5))
  }
  p
}
```

```{r}
#| label: fig-distribution
#| fig-width: 11
#| fig-height: 8

pA <- panel(water_month, freq_month, month,    "Fog", "(A) Fog, monthly")
pB <- panel(water_month, freq_month, month,    "Dew", "(B) Dew, monthly")
pC <- panel(water_hour,  freq_hour,  hour_num, "Fog", "(C) Fog, hourly", hourly = TRUE)
pD <- panel(water_hour,  freq_hour,  hour_num, "Dew", "(D) Dew, hourly", hourly = TRUE)

fig_distribution <- (pA | pB) / (pC | pD) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

fig_distribution

ggsave(file.path(dir_fig, "fig04_seasonal_diurnal_distribution.png"),
       fig_distribution, width = 11, height = 8, dpi = 300, bg = "white")
```

# Values used in the text of Section 3.1

```{r}
#| label: values

freq_month %>% select(month, key, value) %>%
  pivot_wider(names_from = key, values_from = value) %>% arrange(month) %>% print(n = Inf)
freq_hour %>% select(hour_num, key, value) %>%
  pivot_wider(names_from = key, values_from = value) %>% arrange(hour_num) %>% print(n = Inf)
water_month %>% select(month, key, value) %>%
  pivot_wider(names_from = key, values_from = value) %>% arrange(month) %>% print(n = Inf)
water_hour %>% select(hour_num, key, value) %>%
  pivot_wider(names_from = key, values_from = value) %>% arrange(hour_num) %>% print(n = Inf)
```
