---
title: "Figures of Section 3.2: vertical profiles and gradients under fog, dew and no occurrence"
author: "Francisca Muñoz Narbona"
format:
  html:
    toc: true
    self-contained: true
editor: visual
editor_options:
  chunk_output_type: console
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
    dpi: 300
---

Figures 5 and 6 of the manuscript, from `data/derived/stations_theta_q_2024.csv` (step 4).

- Figure 5 (`figures/fig05_profiles_typical_days.png`): vertical profiles of θ, q and U along the transect on 28 April 2024 (fog) and 24 April 2024 (dew), the same days as Figure 3. Hourly records (minute 00). For fog and dew, the most contrasting hour in θ between both regimes is shown (the fog hour with the smallest θ increase between 48 and 1211 m and the dew hour with the largest increase); for no occurrence, the first (solid line) and last (dashed line) hour of the day at OYA1211.
- Figure 6 (`figures/fig06_vertical_gradients.png`): mean vertical gradients (∂q/∂z, ∂θ/∂z, ∂U/∂z) between the coastal reference station (48 m) and each station, for fog, dew and no occurrence (mean ± 1 SD of the 10-min records). Stations are ordered by elevation with constant spacing (not to scale) and OYA1211, the anchor station of the classification, is highlighted. The 48 m point is shown with zero gradient (it is the reference). OYA1069 is omitted (no valid temperature or wind data in 2024). The ∂θ/∂z axis is limited to -0.2 to 9.95 x 10⁻³ K m⁻¹ to widen the separation between fog and dew; all means remain visible but the no-occurrence standard deviations are cut. Dashed vertical lines: thresholds of Lobos-Roco et al. (2018) (dark red) and of this study (blue fog, orange dew, grey ∂q/∂z).

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)

dir_data <- "../data"
dir_fig  <- "../figures"
```

```{r}
#| label: data-gradients

# Output of 04_thermodynamic_variables.qmd: pressure (measured where available,
# otherwise hypsometric from the coastal reference), θ referred to the pressure of
# the coastal reference and q (Magnus), with the OYA1211 classification attached
dat <- read_csv(file.path(dir_data, "derived/stations_theta_q_2024.csv"), show_col_types = FALSE) %>%
  mutate(datetime = as.POSIXct(datetime, tz = "UTC"))

# Gradients relative to the coastal reference (airport, 48 m), every 10 min
grad <- dat %>%
  group_by(datetime) %>%
  mutate(
    theta_ref = theta_K[estacion == "AEROPUERTO"][1],
    q_ref     = q_gkg[estacion == "AEROPUERTO"][1],
    u_ref     = viento_vel_m_s[estacion == "AEROPUERTO"][1]
  ) %>%
  ungroup() %>%
  mutate(
    grad_theta = (theta_K - theta_ref) / (z_m - 48),
    grad_q     = (q_gkg - q_ref) / (z_m - 48),
    grad_u     = (viento_vel_m_s - u_ref) / (z_m - 48)
  ) %>%
  filter(z_m != 48, estacion != "OYA_1069",
         clasif_evento %in% c("NIEBLA", "ROCIO", "SIN_EVENTO"))

gradient_summary <- grad %>%
  pivot_longer(c(grad_q, grad_theta, grad_u), names_to = "variable", values_to = "value") %>%
  group_by(variable, clasif_evento, z_m) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), sd = sd(value, na.rm = TRUE),
            median_value = median(value, na.rm = TRUE), n = sum(!is.na(value)), .groups = "drop") %>%
  # reference point at 48 m: zero gradient by definition (no standard deviation)
  bind_rows(expand_grid(variable = c("grad_q", "grad_theta", "grad_u"),
                        clasif_evento = c("NIEBLA", "ROCIO", "SIN_EVENTO"),
                        z_m = 48, mean_value = 0, sd = NA_real_, median_value = 0, n = NA_integer_)) %>%
  mutate(
    regime = factor(recode(clasif_evento, "NIEBLA" = "Fog", "ROCIO" = "Dew",
                            "SIN_EVENTO" = "No occurrence"),
                     levels = c("Fog", "Dew", "No occurrence")),
    station_lab = factor(paste0(z_m, " m"), levels = paste0(sort(unique(z_m)), " m")),
    y = as.numeric(station_lab) + case_when(regime == "Fog" ~ 0.2,
                                             regime == "Dew" ~ 0,
                                             TRUE ~ -0.2)
  ) %>%
  arrange(variable, regime, z_m)   # ordered by elevation so that geom_path joins the points correctly
```

# Figure 6: mean vertical gradients

```{r}
#| label: fig-gradients

col_regime <- c("Fog" = "#4292c6", "Dew" = "#c97a4a", "No occurrence" = "#7f7f7f")
levels_y <- levels(gradient_summary$station_lab)
y_1211 <- which(levels_y == "1211 m")

theme_grad <- theme_classic(base_size = 10) +
  theme(
    legend.title    = element_blank(),
    legend.position = "bottom",
    axis.line       = element_line(color = "#c3c2b7"),
    axis.ticks      = element_line(color = "#c3c2b7"),
    panel.grid.major.x = element_line(color = "#ecebe6", linewidth = 0.3),
    plot.title      = element_text(face = "bold", size = 10)
  )

# reference thresholds (in units of 10^-3)
thresholds <- tribble(
  ~variable,    ~x,    ~color,    ~label, ~level,
  "grad_theta",  2.6,  "#4292c6", "2.6",     1,
  "grad_theta",  3.1,  "#8b1a1a", "3.1",     2,
  "grad_theta",  3.4,  "#c97a4a", "3.4",     3,
  "grad_q",     -1.6,  "#8b1a1a", "-1.6",    1,
  "grad_q",     -0.4,  "#525252", "-0.4",    1
)
# Lobos-Roco et al. (2018): dark red; this study: blue (fog), orange (dew), grey (Δq/Δz)

panel_grad <- function(var, title, xlab, xlim = NULL) {
  d <- gradient_summary %>% filter(variable == var)
  u <- thresholds %>% filter(variable == var)
  p <- ggplot(d, aes(y = y, color = regime)) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = y_1211 - 0.45, ymax = y_1211 + 0.45,
             fill = "#f3f1ea", color = NA) +
    geom_vline(xintercept = 0, color = "#c3c2b7", linewidth = 0.4)
  if (nrow(u) > 0) {
    p <- p +
      geom_vline(data = u, aes(xintercept = x), color = u$color,
                 linetype = "22", linewidth = 0.45, inherit.aes = FALSE)
  }
  p <- p +
    geom_segment(aes(x = (mean_value - sd) * 1000, xend = (mean_value + sd) * 1000, yend = y),
                 linewidth = 1.1, alpha = 0.2, lineend = "round") +
    geom_path(aes(x = mean_value * 1000, group = regime), linewidth = 0.7) +
    geom_point(aes(x = mean_value * 1000), size = 2) +
    scale_color_manual(values = col_regime) +
    scale_y_continuous(breaks = seq_along(levels_y), labels = levels_y,
                       expand = expansion(add = c(0.5, 0.5))) +
    coord_cartesian(xlim = xlim, ylim = c(0.6, length(levels_y) + 1.75), clip = "off") +
    labs(title = title, x = xlab, y = NULL) +
    theme_grad
  if (nrow(u) > 0) {
    p <- p + geom_label(data = u, aes(x = x, y = length(levels_y) + 0.35 + 0.42 * level, label = label),
                        color = u$color, fill = "white", label.size = 0, label.padding = unit(0.08, "lines"),
                        size = 2.6, fontface = "bold", inherit.aes = FALSE)
  }
  p
}

pq  <- panel_grad("grad_q",     "(A) Specific humidity",  expression(Delta*q/Delta*z~(10^{-3}~g~kg^{-1}~m^{-1}))) +
  labs(y = "Station elevation (m a.s.l.)")
# Δθ/Δz: X axis limited to widen the separation between fog and dew; the
# no-occurrence standard deviations (up to ~14 x 10^-3) are cut, but all the
# means remain visible.
pth <- panel_grad("grad_theta", "(B) Potential temperature", expression(Delta*theta/Delta*z~(10^{-3}~K~m^{-1})),
                  xlim = c(-0.2, 9.95)) +
  scale_x_continuous(breaks = seq(0, 8, 2), expand = expansion(add = 0))
pu  <- panel_grad("grad_u",     "(C) Wind speed",         expression(Delta*U/Delta*z~(10^{-3}~s^{-1})))

# the three panels with the same width
fig_gradients <- (pq | pth | pu) +
  plot_layout(guides = "collect", widths = c(1, 1, 1)) &
  theme(legend.position = "bottom")

fig_gradients

ggsave(file.path(dir_fig, "fig06_vertical_gradients.png"),
       fig_gradients, width = 12, height = 5.4, dpi = 300, bg = "white")
```

```{r}
#| label: values
# Values used in the text of Section 3.2
gradient_summary %>% select(variable, regime, z_m, mean_value, sd, median_value, n) %>%
  arrange(variable, regime, z_m) %>% print(n = Inf)
```

# Figure 5: vertical profiles on typical days

```{r}
#| label: fig-profiles-days
#| fig-width: 10
#| fig-height: 6.5

days_sel <- tribble(
  ~date,            ~day_label,
  ymd("2024-04-28"), "28 April 2024 (fog)",
  ymd("2024-04-24"), "24 April 2024 (dew)"
)

dat_hours <- dat %>%
  filter(minute(datetime) == 0, estacion != "OYA_1069") %>%
  mutate(date = as.Date(datetime), hour = hour(datetime))

# No-occurrence hours: first and last of the day at OYA1211
hours_none <- dat_hours %>%
  filter(date %in% days_sel$date, estacion == "OYA_1211", clasif_evento == "SIN_EVENTO") %>%
  group_by(date, clasif_evento) %>%
  summarise(first = min(hour), last = max(hour), .groups = "drop") %>%
  pivot_longer(c(first, last), names_to = "moment", values_to = "hour")

# Fog and dew hours: the most contrasting hour in θ between both regimes, i.e.
# the fog hour with the smallest θ increase between 48 and 1211 m (most
# homogeneous profile) and the dew hour with the largest increase (most stratified).
dtheta_hour <- dat_hours %>%
  filter(date %in% days_sel$date) %>%
  group_by(date, hour) %>%
  summarise(dtheta = theta_K[z_m == 1211][1] - theta_K[z_m == 48][1], .groups = "drop") %>%
  left_join(dat_hours %>% filter(estacion == "OYA_1211") %>% select(date, hour, clasif_evento),
            by = c("date", "hour"))
print(dtheta_hour %>% filter(clasif_evento %in% c("NIEBLA", "ROCIO")))

hours_occ <- bind_rows(
  dtheta_hour %>% filter(clasif_evento == "NIEBLA") %>% slice_min(dtheta, n = 1, with_ties = FALSE),
  dtheta_hour %>% filter(clasif_evento == "ROCIO")  %>% slice_max(dtheta, n = 1, with_ties = FALSE)
) %>%
  transmute(date, clasif_evento, moment = "occurrence", hour)

hours_sel <- bind_rows(hours_occ, hours_none)

profiles <- dat_hours %>%
  inner_join(hours_sel, by = c("date", "hour", "clasif_evento")) %>%
  inner_join(days_sel, by = "date") %>%
  mutate(
    day_label = factor(day_label, levels = days_sel$day_label),
    series_leg = case_when(
      clasif_evento == "NIEBLA" ~ "Fog",
      clasif_evento == "ROCIO"  ~ "Dew",
      moment == "first"      ~ "No occurrence, first hour",
      TRUE                      ~ "No occurrence, last hour"
    ),
    series_leg = factor(series_leg, levels = c("Fog", "Dew", "No occurrence, first hour",
                                             "No occurrence, last hour")),
    series   = paste(series_leg, hour),
    station_lab = factor(paste0(z_m, " m"), levels = paste0(sort(unique(z_m)), " m")),
    y = as.numeric(station_lab)
  ) %>%
  select(day_label, series_leg, hour, series, z_m, station_lab, y,
         theta_K, q_gkg, viento_vel_m_s) %>%
  pivot_longer(c(theta_K, q_gkg, viento_vel_m_s), names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("theta_K", "q_gkg", "viento_vel_m_s"),
                           labels = c("Potential~temperature~(K)",
                                      "Specific~humidity~(g~kg^{-1})",
                                      "Wind~speed~(m~s^{-1})"))) %>%
  arrange(day_label, series, variable, z_m)

# hours shown (for the caption)
print(hours_sel %>% arrange(date, clasif_evento, hour))

levels_y_d <- levels(profiles$station_lab)
y_1211_d <- which(levels_y_d == "1211 m")

fig_profiles_days <- ggplot(profiles, aes(x = value, y = y, color = series_leg,
                                      linetype = series_leg, shape = series_leg, group = series)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = y_1211_d - 0.45, ymax = y_1211_d + 0.45,
           fill = "#f3f1ea", color = NA) +
  geom_path(linewidth = 0.6, na.rm = TRUE) +
  geom_point(size = 1.8, na.rm = TRUE) +
  facet_grid(day_label ~ variable, scales = "free_x",
             labeller = labeller(variable = label_parsed, day_label = label_value)) +
  scale_color_manual(values = c("Fog" = "#4292c6", "Dew" = "#c97a4a",
                                "No occurrence, first hour" = "#7f7f7f",
                                "No occurrence, last hour" = "#7f7f7f"), drop = FALSE) +
  scale_linetype_manual(values = c("Fog" = "solid", "Dew" = "solid",
                                   "No occurrence, first hour" = "solid",
                                   "No occurrence, last hour" = "22"), drop = FALSE) +
  scale_shape_manual(values = c("Fog" = 16, "Dew" = 16,
                                "No occurrence, first hour" = 16,
                                "No occurrence, last hour" = 1), drop = FALSE) +
  scale_y_continuous(breaks = seq_along(levels_y_d), labels = levels_y_d,
                     expand = expansion(add = 0.5)) +
  labs(x = NULL, y = "Station elevation (m a.s.l.)") +
  theme_classic(base_size = 10) +
  theme(
    legend.title     = element_blank(),
    legend.position  = "bottom",
    axis.line        = element_line(color = "#c3c2b7"),
    axis.ticks       = element_line(color = "#c3c2b7"),
    panel.grid.major.x = element_line(color = "#ecebe6", linewidth = 0.3),
    panel.border     = element_rect(color = "#c3c2b7", fill = NA, linewidth = 0.4),
    strip.background = element_rect(fill = "#f9f9f7", color = NA),
    strip.text       = element_text(face = "bold"),
    strip.text.y     = element_text(angle = -90),
    panel.spacing.x  = unit(1.2, "lines")
  )

fig_profiles_days

ggsave(file.path(dir_fig, "fig05_profiles_typical_days.png"),
       fig_profiles_days, width = 10, height = 6.5, dpi = 300, bg = "white")
```
