---
title: "Figures of Section 3.1: typical fog and dew days at OYA1211"
author: "Francisca Muñoz Narbona"
format:
  html:
    toc: true
    self-contained: true
editor: visual
editor_options:
  chunk_output_type: console
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
    dpi: 300
---

Figure 3 of the manuscript (typical fog and dew days) and Figure A1 of the appendix (a day without occurrence and two particular cases). Each panel shows, for one day at OYA1211, the visibility (blue line; red dashed line, 1000 m threshold), the final classification (fog, dew, no occurrence), the instrumental collection (fog collector, dew collector or both) and the GOES-16 fog and low clouds signal.

Input: `data/derived/oya1211_classification_2024.csv` (step 3). Outputs: `figures/fig03_typical_days.png` and `figures/figA1_particular_cases.png`.

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)
library(cowplot)

dir_data <- "../data"
dir_fig  <- "../figures"
```

```{r}
#| label: data-days

# Final classification of OYA1211 (output of 03_classify_fog_dew.qmd)
oya1211_class <- read_csv(
  file.path(dir_data, "derived/oya1211_classification_2024.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    ts_10 = as.POSIXct(ts_10, tz = "UTC"),
    fecha = as.Date(fecha)
  )

fog_days <- c("2024-04-28", "2024-09-21")
dew_days  <- c("2024-04-24", "2024-11-28")
other_days  <- c("2024-08-11", "2024-05-17", "2024-06-27")

selected_days <- tribble(
  ~fecha,            ~label,
  ymd("2024-04-28"), "28 April (fog)",
  ymd("2024-09-21"), "21 September (fog)",
  ymd("2024-04-24"), "24 April (dew)",
  ymd("2024-11-28"), "28 November (dew)",
  ymd("2024-08-11"), "11 August (no event)",
  ymd("2024-05-17"), "17 May (particular case)",
  ymd("2024-06-27"), "27 June (particular case)"
)

day_labels <- setNames(selected_days$label, as.character(selected_days$fecha))

# Instrumental collection (before the classification), from fog_cap/dew_cap
col_collection <- c(
  "Fog"      = "#93C5FD",
  "Dew"      = "#FCA5A5",
  "Both"     = "#756bb1",
  "No event" = "#bdbdbd"
)

dsub <- oya1211_class %>%
  inner_join(selected_days, by = "fecha") %>%
  mutate(
    fecha = as.character(fecha),
    class_fac = factor(
      recode(clasif_evento, "NIEBLA" = "Fog", "ROCIO" = "Dew", "SIN_EVENTO" = "No event"),
      levels = c("Fog", "Dew", "No event")
    ),
    collection_fac = case_when(
      fog_cap == 1 & dew_cap == 1 ~ "Both",
      fog_cap == 1 & dew_cap == 0 ~ "Fog",
      fog_cap == 0 & dew_cap == 1 ~ "Dew",
      TRUE                        ~ "No event"
    ),
    collection_fac = factor(collection_fac, levels = names(col_collection)),
    collection_color = col_collection[as.character(collection_fac)]
  )

band_h <- 500
y_cls  <- -500
y_inst <- -1200
y_goes <- -1900

plot_day_row <- function(df, dates_group, ncol_group, title_group, show_x = FALSE) {

  d <- df %>%
    filter(fecha %in% dates_group) %>%
    mutate(fecha = factor(fecha, levels = dates_group))

  ggplot(d, aes(x = ts_10)) +
    # GOES
    geom_tile(aes(y = y_goes, alpha = fog_sat), fill = "grey30",
              height = band_h, width = 600) +
    scale_alpha_continuous(range = c(0.15, 0.9), na.value = 0, guide = "none") +
    # Instrumental collection: one layer per category with a fixed colour
    # (outside aes), so it does not interfere with the "Classification" scale
    lapply(names(col_collection), function(k) {
      geom_tile(data = filter(d, collection_fac == k), aes(y = y_inst),
                fill = col_collection[[k]], height = band_h, width = 600)
    }) +
    # Classification
    geom_tile(aes(y = y_cls, fill = class_fac),
              height = band_h, width = 600) +
    scale_fill_manual(
      values = c("Fog" = "#93C5FD", "Dew" = "#FCA5A5", "No event" = "#bdbdbd"),
      name = "Classification",
      drop = FALSE
    ) +
    # Visibility
    geom_line(aes(y = vis_m, group = fecha), color = "#08519c",
              linewidth = 0.45, na.rm = TRUE) +
    geom_hline(yintercept = 1000, linetype = "dashed", color = "red", linewidth = 0.4) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    scale_y_continuous(
      limits = c(-2400, 5000),
      breaks = c(-1900, -1200, -500, 0, 1000, 2000, 3000, 4000, 5000),
      labels = c("GOES", "Instr.", "Classif.", "0", "1000", "2000", "3000", "4000", "5000"),
      expand = expansion(mult = c(0.01, 0.03))
    ) +
    scale_x_datetime(date_breaks = "3 hours", date_labels = "%H:%M") +
    facet_wrap(~ fecha, ncol = ncol_group, scales = "free_x",
               labeller = as_labeller(day_labels)) +
    labs(
      x = if (show_x) "Time (UTC)" else NULL,
      y = "Visibility (m)",
      title = title_group
    ) +
    theme_bw(base_size = 11) +
    theme(
      legend.position = "bottom",
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      axis.text.y = element_text(face = "bold"),
      plot.title = element_text(face = "bold", size = 11),
      plot.margin = margin(4, 6, 4, 6)
    )
}

# Shared legends (classification + instrumental collection)
p_legend_class <- ggplot(
  tibble(k = factor(c("Fog", "Dew", "No event"), levels = c("Fog", "Dew", "No event")), x = 1, y = 1),
  aes(x, y, fill = k)
) +
  geom_tile() +
  scale_fill_manual(values = c("Fog" = "#93C5FD", "Dew" = "#FCA5A5", "No event" = "#bdbdbd"),
                    name = "Classification", drop = FALSE) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5)) +
  theme(legend.position = "bottom")

p_legend_collection <- ggplot(
  tibble(k = factor(names(col_collection), levels = names(col_collection)), x = 1, y = 1),
  aes(x, y, fill = k)
) +
  geom_tile() +
  scale_fill_manual(values = col_collection, name = "Instrumental collection", drop = FALSE) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5)) +
  theme(legend.position = "bottom")

legend_row <- wrap_elements(cowplot::get_legend(p_legend_class)) |
  wrap_elements(cowplot::get_legend(p_legend_collection))
```

```{r}
#| label: fig-typical-days
#| fig-width: 10
#| fig-height: 8

p_fog <- plot_day_row(dsub, fog_days, ncol_group = 2, title_group = "Fog")
p_dew  <- plot_day_row(dsub, dew_days,  ncol_group = 2, title_group = "Dew", show_x = TRUE)

fig_typical_days <- wrap_plots(
  p_fog + theme(legend.position = "none"),
  p_dew  + theme(legend.position = "none"),
  legend_row,
  ncol = 1,
  heights = c(1, 1, 0.15)
)

fig_typical_days

ggsave(file.path(dir_fig, "fig03_typical_days.png"),
       fig_typical_days, width = 10, height = 8, dpi = 300, bg = "white")
```

```{r}
#| label: fig-particular-cases
#| fig-width: 10
#| fig-height: 5

p_other <- plot_day_row(dsub, other_days, ncol_group = 3,
                          title_group = "No event and particular cases", show_x = TRUE)

fig_particular_cases <- wrap_plots(
  p_other + theme(legend.position = "none"),
  legend_row,
  ncol = 1,
  heights = c(1, 0.22)
)

fig_particular_cases

ggsave(file.path(dir_fig, "figA1_particular_cases.png"),
       fig_particular_cases, width = 10, height = 5, dpi = 300, bg = "white")
```
